Type Safe Redis Queries

### 2017-07-13 TyDe2017/sections/TypeLevelDict-orig.tex
Auto stash before merge of "master" and "origin/master"

One of the challenges of statically ensuring type correctness of \Redis{},
which also presents in other stateful languages, is that the type of the value
associated to a key can be altered after updating. To ensure type correctness,
we keep track of the types of all existing keys in a {\em dictionary} ---
conceptually, an associate list, or a list of pairs of keys and \Redis{} types.
For example, the dictionary \ensuremath{[\mskip1.5mu (\text{\tt \char34 A\char34},\Conid{Int}),(\text{\tt \char34 B\char34},\Conid{Char}),(\text{\tt \char34 C\char34},\Conid{Bool})\mskip1.5mu]} represents
a predicate stating that ``the keys \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}, and \ensuremath{\text{\tt \char34 C\char34}} are respectively
associated to values of type \ensuremath{\Conid{Int}}, \ensuremath{\Conid{Char}}, and \ensuremath{\Conid{Bool}}.''

The dictionary mixes values (strings such as \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}) and types. Further
more, as mentioned in Section~\ref{sec:indexed-monads}, the dictionaries are
to be used parameters to the indexed monad \ensuremath{\Conid{Edis}}. In a dependently typed
programming language (without the so-called ``phase distinction'' ---
separation between types and terms), this would pose no problem. In Haskell
however, the dictionaries, to index a monad, has to be a type as well.

@@

Haskell maintains the distinction between values, types, and kinds: values are
categorized by types, and types are categorized by kinds. The kinds are relatively simple: \ensuremath{\mathbin{*}} is the kind of all {\em lifted} types, while type
constructors have kinds such as \ensuremath{\mathbin{*}\to \mathbin{*}}, \ensuremath{\mathbin{*}\to \mathbin{*}\to \mathbin{*}}, etc. Consider the
datatype definitions below:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{Nat}\mathrel{=}\Conid{Zero}\mid \Conid{Suc}\;\Conid{Nat}~~,\qquad\;\mathbf{data}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\mathrel{=}[\mskip1.5mu \mskip1.5mu]\mid \Varid{a}\mathbin{:}[\mskip1.5mu \Varid{a}\mskip1.5mu]~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The lefthand side is usually seen as having defined a type \ensuremath{\Conid{Nat}\mathbin{::}\mathbin{*}},
and two value constructors \ensuremath{\Conid{Zero}\mathbin{::}\Conid{Nat}} and \ensuremath{\Conid{Suc}\mathbin{::}\Conid{Nat}\to \Conid{Nat}}. The righthand
side is how Haskell lists are understood. The {\em kind} of \ensuremath{[\mskip1.5mu \mathbin{\cdot}\mskip1.5mu]} is \ensuremath{\mathbin{*}\to \mathbin{*}},
since it takes a lifted type \ensuremath{\Varid{a}} to a lifted type \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]}. The two value constructors respectively have types \ensuremath{[\mskip1.5mu \mskip1.5mu]\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]} and \ensuremath{(\mathbin{:})\mathbin{::}\Varid{a}\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]}, for all type \ensuremath{\Varid{a}}.

The GHC extension \emph{data kinds}~\cite{promotion}, however, automatically
promotes certain ``suitable'' types to kinds.\footnote{It is only informally
described in the GHC manual what types are ``suitable''.} With the extension,
the \ensuremath{\mathbf{data}} definitions above has an alternative reading: \ensuremath{\Conid{Nat}} is a new kind,
\ensuremath{\Conid{Zero}\mathbin{::}\Conid{Nat}} is a type having kind \ensuremath{\Conid{Nat}}, and \ensuremath{\Conid{Suc}\mathbin{::}\Conid{Nat}\to \Conid{Nat}} is a type
constructor, taking a type in kind \ensuremath{\Conid{Nat}} to another type in \ensuremath{\Conid{Nat}}. Whether a
constructor is promoted can often be inferred from the context. To be more specific, prefixing a constructor with a single quote, such as in \ensuremath{\mbox{\textquotesingle}\Conid{Zero}} and
\ensuremath{\mbox{\textquotesingle}\Conid{Suc}}, denotes that it is promoted.

The situation of lists is similar: for all kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]} is also a kind. For
all kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \mskip1.5mu]\mathbin{::}[\mskip1.5mu \Varid{k}\mskip1.5mu]} is a type. Given a type \ensuremath{\Varid{x}}
of kind \ensuremath{\Varid{k}} and a type \ensuremath{\Varid{xs}} of kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}, \ensuremath{\Varid{x}\mathbin{:}\Varid{xs}} is again a type of
kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}. Formally, \ensuremath{(\mathbin{:})\mathbin{::}\Varid{k}\to [\mskip1.5mu \Varid{k}\mskip1.5mu]\to [\mskip1.5mu \Varid{k}\mskip1.5mu]}. For example,
\ensuremath{\Conid{Int}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Char}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Bool}\mathbin{\mbox{\textquotesingle}\!:}\mbox{\textquotesingle}[~]))} is a type having kind \ensuremath{[\mskip1.5mu \mathbin{*}\mskip1.5mu]} --- a list of
(lifted) types. The optional quote denotes that the constructors are promoted.
The same list can be denoted by a syntax sugar \ensuremath{\mbox{\textquotesingle}[\Conid{Int},\Conid{Char},\Conid{Bool}]}.

@@

Strings in Haskell are nothing but lists of \ensuremath{\Conid{Char}}s. Regarding promotion,
however, a \ensuremath{\Conid{String}} can be promoted to a type having kind \ensuremath{\Conid{Symbol}}. \ensuremath{\Conid{Symbol}} is
a type without a constructor: \ensuremath{\mathbf{data}\;\Conid{Symbol}},
intended to be used as a promoted kind. In the expression:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\text{\tt \char34 this~is~a~type-level~string~literal\char34}\mathbin{::}\Conid{Symbol}~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
the string on the lefthand side of \ensuremath{(\mathbin{::})} is a type, having kind \ensuremath{\Conid{Symbol}}.

With all of these ingredients, we are ready to build our dictionaries, or
type-level associate lists:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Conid{DictEmpty}\mathrel{=}\mbox{\textquotesingle}[~]~~,{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Conid{Dict0}\mathrel{=}\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 key\char34},\Conid{Bool})]~~,{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Conid{Dict1}\mathrel{=}\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{Int}),\mbox{\textquotesingle}(\text{\tt \char34 B\char34},\text{\tt \char34 A\char34})]~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
All the entities defined above are types, where \ensuremath{\Conid{Dict0}} and \ensuremath{\Conid{Dict1}}
have kind \ensuremath{[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]}.

\subsection{Type-Level Functions}

Now that we can represent dictionaries as types, the next step is to define
operations on them. A function that inserts an entry to a dictionary, for
example, is a function from a type to a type. While it was shown that it is
possible to simulate type-level functions using Haskell type
classes~\cite{McBride:02:Faking}, in recent versions of GHC, {\em indexed type
families}, or type families for short, are considered a cleaner solution.

For example, compare conjunction \ensuremath{(\mathrel{\wedge})} and its type-level
counterpart \ensuremath{\Conid{And}}:\\
\noindent{\centering %\small
\begin{minipage}[b]{0.35\linewidth}
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}c<{\hspost}@{}}%
\column{7E}{@{}l@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\mathrel{\wedge})\mathbin{::}\Conid{Bool}\to \Conid{Bool}\to \Conid{Bool}{}\<[E]%
\\
\>[B]{}\Conid{True}{}\<[7]%
\>[7]{}\mathrel{\wedge}{}\<[7E]%
\>[11]{}\Conid{True}{}\<[17]%
\>[17]{}\mathrel{=}\Conid{True}{}\<[E]%
\\
\>[B]{}\Varid{a}{}\<[7]%
\>[7]{}\mathrel{\wedge}{}\<[7E]%
\>[11]{}\Varid{b}{}\<[17]%
\>[17]{}\mathrel{=}\Conid{False}~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\end{minipage}
\begin{minipage}[b]{0.55\linewidth}
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{27}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{And}\;(\Varid{a}\mathbin{::}\Conid{Bool})\;(\Varid{b}\mathbin{::}\Conid{Bool})\mathbin{::}\Conid{Bool}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathbf{where}\;{}\<[10]%
\>[10]{}\Conid{And}\;{}\<[15]%
\>[15]{}\Conid{True}\;{}\<[21]%
\>[21]{}\Conid{True}{}\<[27]%
\>[27]{}\mathrel{=}\Conid{True}{}\<[E]%
\\
\>[10]{}\Conid{And}\;{}\<[15]%
\>[15]{}\Varid{a}\;{}\<[21]%
\>[21]{}\Varid{b}{}\<[27]%
\>[27]{}\mathrel{=}\Conid{False}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\end{minipage}
}\\
The lefthand side is a typical definition of \ensuremath{(\mathrel{\wedge})} by pattern matching.
On the righthand side, \ensuremath{\Conid{Bool}} is not a type, but a type lifted to a kind,
while \ensuremath{\Conid{True}} and \ensuremath{\Conid{False}} are types of kind \ensuremath{\Conid{Bool}}. The declaration says
that \ensuremath{\Conid{And}} is a family of types, indexed by two parameters \ensuremath{\Varid{a}} and \ensuremath{\Varid{b}} of
kind \ensuremath{\Conid{Bool}}. The type with index \ensuremath{\Conid{True}} and \ensuremath{\Conid{True}} is \ensuremath{\Conid{True}}, and all
other indices lead to \ensuremath{\Conid{False}}. For our purpose, we can read \ensuremath{\Conid{And}} as a type-level function. Observe how it resembles the term-level \ensuremath{(\mathrel{\wedge})}.

Note that type families in Haskell come in many flavors. Families can be
defined for \ensuremath{\mathbf{data}} and \ensuremath{\mathbf{type}} synonym. They can appear inside type
classes~\cite{tfclass,tfsynonym} or at toplevel. Toplevel type families
can be open~\cite{tfopen} or closed~\cite{tfclosed}. The flavor we chose
is top-level, closed type synonym family, since it allows overlapping
instances, and we need none of the extensibility provided by open type
families. Notice that the instance \ensuremath{\Conid{And}\;\Conid{True}\;\Conid{True}} could be subsumed under
the more general instance, \ensuremath{\Conid{And}\;\Varid{a}\;\Varid{b}}. In a closed type family we may resolve
the overlapping in order, just like how cases overlapping is resolved in term-level functions.

%\subsection{Functions on Type-Level Dictionaries}

We are now able to define operations on type-level dictionaries.
Let's begin with dictionary lookup.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{32}{@{}>{\hspre}c<{\hspost}@{}}%
\column{32E}{@{}l@{}}%
\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Get}\;(\Varid{xs}\mathbin{::}[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu])\;(\Varid{s}\mathbin{::}\Conid{Symbol})\mathbin{::}\mathbin{*}\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Get}\;(\mbox{\textquotesingle}(\Varid{s},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{s}{}\<[32]%
\>[32]{}\mathrel{=}{}\<[32E]%
\>[35]{}\Varid{x}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Get}\;(\mbox{\textquotesingle}(\Varid{t},\Varid{x})\mathbin{\mbox{\textquotesingle}\!:}\Varid{xs})\;\Varid{s}{}\<[32]%
\>[32]{}\mathrel{=}{}\<[32E]%
\>[35]{}\Conid{Get}\;\Varid{xs}\;\Varid{s}~~{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The type-level function \ensuremath{\Conid{Get}} returns the entry associated with key \ensuremath{\Varid{s}} in the
dictionary \ensuremath{\Varid{xs}}. Notice, in the first case, how type-level equality can be
expressed by unifying type variables with the same name.
Note also that \ensuremath{\Conid{Get}} is a partial function on types:
\ensuremath{\Conid{Get}\;\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{Int})]\;\text{\tt \char34 A\char34}} evaluates to \ensuremath{\Conid{Int}}, but
\ensuremath{\Conid{Get}\;\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{Int})]\;\text{\tt \char34 B\char34}} gets stuck.
... and these types are computed at compile-time. It wouldn't make
much sense for a type checker to crash and throw a ``Non-exhaustive'' error or
be non-terminating.
\todo{So, what exactly happens when we do the second?}

>>>>

One of the challenges of statically ensuring type correctness of stateful
languages is that the type of the value of a key can be altered by updating.
In \Redis{}, one may delete an existing key and create it again by assigning to
it a value of a different type. To ensure type correctness, we keep track of the
types of all existing keys in a {\em dictionary} --- an associate list, or a
list of pairs of keys and some encoding of types. For example, we may use the
dictionary \ensuremath{[\mskip1.5mu (\text{\tt \char34 A\char34},\Conid{Int}),(\text{\tt \char34 B\char34},\Conid{Char}),(\text{\tt \char34 C\char34},\Conid{Bool})\mskip1.5mu]} to represent a
predicate, or a constraint, stating that ``the keys in the data store are \ensuremath{\text{\tt \char34 A\char34}},
\ensuremath{\text{\tt \char34 B\char34}}, and \ensuremath{\text{\tt \char34 C\char34}}, respectively assigned values of type \ensuremath{\Conid{Int}}, \ensuremath{\Conid{Char}}, and
\ensuremath{\Conid{Bool}}.'' (This representation will be refined in the next section.)

The dictionary above mixes values (strings such as \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}) and types.
Furthermore, as mentioned in Section~\ref{sec:indexed-monads}, the
dictionaries will be parameters to the indexed monad \ensuremath{\Conid{Edis}}. In a dependently
typed programming language (without the so-called ``phase distinction'' ---
separation between types and terms), this would pose no problem. In Haskell
however, the dictionaries, to index a monad, has to be a type as well.

@@

Haskell maintains the distinction between values, types, and kinds: values are
categorized by types, and types are categorized by kinds. The kinds are
relatively simple: \ensuremath{\mathbin{*}} is the kind of all {\em lifted} types, while type
constructors have kinds such as \ensuremath{\mathbin{*}\to \mathbin{*}}, or \ensuremath{\mathbin{*}\to \mathbin{*}\to \mathbin{*}}, etc.\footnotemark\,
Consider the datatype definitions below:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{data}\;\Conid{Nat}\mathrel{=}\Conid{Zero}\mid \Conid{Suc}\;\Conid{Nat}~~,\qquad\;\mathbf{data}\;[\mskip1.5mu \Varid{a}\mskip1.5mu]\mathrel{=}[\mskip1.5mu \mskip1.5mu]\mid \Varid{a}\mathbin{:}[\mskip1.5mu \Varid{a}\mskip1.5mu]~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The left-hand side is usually seen as having defined a type \ensuremath{\Conid{Nat}\mathbin{::}\mathbin{*}},
and two value constructors \ensuremath{\Conid{Zero}\mathbin{::}\Conid{Nat}} and \ensuremath{\Conid{Suc}\mathbin{::}\Conid{Nat}\to \Conid{Nat}}. The right-hand
side is how Haskell lists are understood. The {\em kind} of \ensuremath{[\mskip1.5mu \mathbin{\cdot}\mskip1.5mu]} is \ensuremath{\mathbin{*}\to \mathbin{*}},
since it takes a lifted type \ensuremath{\Varid{a}} to a lifted type \ensuremath{[\mskip1.5mu \Varid{a}\mskip1.5mu]}. The two value
constructors respectively have types \ensuremath{[\mskip1.5mu \mskip1.5mu]\mathbin{::}[\mskip1.5mu \Varid{a}\mskip1.5mu]} and \ensuremath{(\mathbin{:})\mathbin{::}\Varid{a}\to [\mskip1.5mu \Varid{a}\mskip1.5mu]\to [\mskip1.5mu \Varid{a}\mskip1.5mu]}, for all type \ensuremath{\Varid{a}}.

\footnotetext{In Haskell, the opposite of \emph{lifted} types are \emph{unboxed}
types, which are not represented by a pointer to a heap object, and cannot be
stored in a polymorphic data type.}

The GHC extension \emph{data kinds}~\cite{promotion}, however, automatically
promotes certain ``suitable'' types to kinds.\footnote{It is only informally
described in the GHC manual what types are ``suitable''.} With the extension,
the \ensuremath{\mathbf{data}} definitions above has an alternative reading: \ensuremath{\Conid{Nat}} is a new kind,
\ensuremath{\Conid{Zero}\mathbin{::}\Conid{Nat}} is a type having kind \ensuremath{\Conid{Nat}}, and \ensuremath{\Conid{Suc}\mathbin{::}\Conid{Nat}\to \Conid{Nat}} is a type
constructor, taking a type in kind \ensuremath{\Conid{Nat}} to another type in \ensuremath{\Conid{Nat}}. When one
sees a constructor in an expression, whether it is promoted can often be
inferred from the context. When one needs to be more specific, prefixing a
constructor with a single quote, such as in \ensuremath{\mbox{\textquotesingle}\Conid{Zero}} and \ensuremath{\mbox{\textquotesingle}\Conid{Suc}}, denotes that it
is promoted.

The situation of lists is similar: for all kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]} is also a kind. For
all kind \ensuremath{\Varid{k}}, \ensuremath{[\mskip1.5mu \mskip1.5mu]} is a type of kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}. Given a type \ensuremath{\Varid{a}} of kind \ensuremath{\Varid{k}} and a
type \ensuremath{\Varid{as}} of kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}, \ensuremath{\Varid{a}\mathbin{:}\Varid{as}} is again a type of kind \ensuremath{[\mskip1.5mu \Varid{k}\mskip1.5mu]}. Formally,
\ensuremath{(\mathbin{:})\mathbin{::}\Varid{k}\to [\mskip1.5mu \Varid{k}\mskip1.5mu]\to [\mskip1.5mu \Varid{k}\mskip1.5mu]}. For example, \ensuremath{\Conid{Int}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Char}\mathbin{\mbox{\textquotesingle}\!:}(\Conid{Bool}\mathbin{\mbox{\textquotesingle}\!:}\mbox{\textquotesingle}[~]))} is a
type having kind \ensuremath{[\mskip1.5mu \mathbin{*}\mskip1.5mu]} --- it is a list of (lifted) types. The optional quote
denotes that the constructors are promoted. The same list can be denoted by a
syntax sugar \ensuremath{\mbox{\textquotesingle}[\Conid{Int},\Conid{Char},\Conid{Bool}]}.

@@

Strings in Haskell are nothing but lists of \ensuremath{\Conid{Char}}s. Regarding promotion,
however, a string can be promoted to a type having kind \ensuremath{\Conid{Symbol}}. In the expression:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\text{\tt \char34 this~is~a~type-level~string~literal\char34}\mathbin{::}\Conid{Symbol}~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
the string on the left-hand side of \ensuremath{(\mathbin{::})} is a type whose kind is \ensuremath{\Conid{Symbol}}.

With all of these ingredients, we are ready to build our dictionaries, or
type-level associate lists:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Conid{DictEmpty}\mathrel{=}\mbox{\textquotesingle}[~]~~,{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Conid{Dict0}\mathrel{=}\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 key\char34},\Conid{Bool})]~~,{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Conid{Dict1}\mathrel{=}\mbox{\textquotesingle}[\mbox{\textquotesingle}(\text{\tt \char34 A\char34},\Conid{Int}),\mbox{\textquotesingle}(\text{\tt \char34 B\char34},\text{\tt \char34 A\char34})]~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
All the entities defined above are types, where \ensuremath{\Conid{Dict0}} has kind \ensuremath{[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]}. In \ensuremath{\Conid{Dict1}}, while \ensuremath{\Conid{Int}} has kind \ensuremath{\mathbin{*}} and \ensuremath{\text{\tt \char34 A\char34}} has kind \ensuremath{\Conid{Symbol}}, the former kind subsumes the later. Thus \ensuremath{\Conid{Dict1}} also has kind \ensuremath{[\mskip1.5mu (\Conid{Symbol},\mathbin{*})\mskip1.5mu]}.

\subsection{Type-Level Functions}
\label{sec:type-fun}

Now that we can represent dictionaries as types, the next step is to define
operations on them. A function that inserts an entry to a dictionary, for
example, is a function from a type to a type. While it was shown that it is
possible to simulate type-level functions using Haskell type
classes~\cite{McBride:02:Faking}, in recent versions of GHC, {\em indexed type
families}, or \emph{type families} for short, are considered a cleaner solution.

For example, compare disjunction \ensuremath{(\mathrel{\vee})} and its type-level
counterpart:\\
\noindent{\centering %\small
\begin{minipage}[b]{0.35\linewidth}
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{7}{@{}>{\hspre}c<{\hspost}@{}}%
\column{7E}{@{}l@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\mathrel{\vee})\mathbin{::}\Conid{Bool}\to \Conid{Bool}\to \Conid{Bool}{}\<[E]%
\\
\>[B]{}\Conid{True}{}\<[7]%
\>[7]{}\mathrel{\vee}{}\<[7E]%
\>[11]{}\Varid{b}{}\<[14]%
\>[14]{}\mathrel{=}\Conid{True}{}\<[E]%
\\
\>[B]{}\Varid{a}{}\<[7]%
\>[7]{}\mathrel{\vee}{}\<[7E]%
\>[11]{}\Varid{b}{}\<[14]%
\>[14]{}\mathrel{=}\Varid{b}~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\end{minipage}
\begin{minipage}[b]{0.55\linewidth}
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{24}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Or}\;(\Varid{a}\mathbin{::}\Conid{Bool})\;(\Varid{b}\mathbin{::}\Conid{Bool})\mathbin{::}\Conid{Bool}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathbf{where}\;{}\<[10]%
\>[10]{}\mbox{\textquotesingle}\Conid{True}{}\<[16]%
\>[16]{}\mathrel{\Conid{`Or`}}\Varid{b}{}\<[24]%
\>[24]{}\mathrel{=}\mbox{\textquotesingle}\Conid{True}{}\<[E]%
\\
\>[10]{}\Varid{a}{}\<[16]%
\>[16]{}\mathrel{\Conid{`Or`}}\Varid{b}{}\<[24]%
\>[24]{}\mathrel{=}\Varid{b}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\end{minipage}
}\\
The left-hand side is a typical definition of \ensuremath{(\mathrel{\vee})} by pattern matching.
On the right-hand side, \ensuremath{\Conid{Bool}} is not a type, but a type lifted to a kind,
while \ensuremath{\Conid{True}} and \ensuremath{\Conid{False}} are types of kind \ensuremath{\Conid{Bool}}. The declaration says that
\ensuremath{\Conid{Or}} is a family of types, indexed by two parameters \ensuremath{\Varid{a}} and \ensuremath{\Varid{b}} of kind \ensuremath{\Conid{Bool}}.
The type with index \ensuremath{\mbox{\textquotesingle}\Conid{True}} and \ensuremath{\Varid{b}} is \ensuremath{\mbox{\textquotesingle}\Conid{True}}, and all other indices lead to \ensuremath{\Varid{b}}.
For our purpose, we can read \ensuremath{\Conid{Or}} as a function from types to types ---
observe how it resembles the term-level \ensuremath{(\mathrel{\vee})}. We present two more type-level
functions about \ensuremath{\Conid{Bool}} --- negation, and conditional, that we will use later:\\
\noindent{\centering %\small
\begin{minipage}[b]{0.35\linewidth}
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{Not}\;\Varid{a}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Conid{Not}\;\mbox{\textquotesingle}\Conid{False}{}\<[14]%
\>[14]{}\mathrel{=}\mbox{\textquotesingle}\Conid{True}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Conid{Not}\;\mbox{\textquotesingle}\Conid{True}{}\<[14]%
\>[14]{}\mathrel{=}\mbox{\textquotesingle}\Conid{False}~~,{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\end{minipage}
\begin{minipage}[b]{0.55\linewidth}
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{12}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Varid{family}\;\Conid{If}\;(\Varid{c}\mathbin{::}\Conid{Bool})\;(\Varid{t}\mathbin{::}\Varid{a})\;(\Varid{f}\mathbin{::}\Varid{a})\mathbin{::}\Varid{a}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Conid{If}\;\mbox{\textquotesingle}\Conid{True}\;{}\<[12]%
\>[12]{}\Varid{tru}\;{}\<[17]%
\>[17]{}\Varid{fls}\mathrel{=}\Varid{tru}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Conid{If}\;\mbox{\textquotesingle}\Conid{False}\;\Varid{tru}\;{}\<[17]%
\>[17]{}\Varid{fls}\mathrel{=}\Varid{fls}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\end{minipage}
}

As a remark, type families in Haskell come in many flavors. One can define families of \ensuremath{\mathbf{data}}, as well as families of \ensuremath{\mathbf{type}} synonym. They can appear
inside type classes~\cite{tfclass,tfsynonym} or at toplevel. Top-level type families can be open~\cite{tfopen} or closed~\cite{tfclosed}. The flavor we
chose is top-level, closed type synonym family, since it allows overlapping
instances, and since we need none of the extensibility provided by open type
families. Notice that the instance \ensuremath{\mbox{\textquotesingle}\Conid{True}\mathrel{\Conid{`Or`}}\Varid{b}} could be subsumed under the
more general instance, \ensuremath{\Varid{a}\mathrel{\Conid{`Or`}}\Varid{b}}. In a closed type family we may resolve the
overlapping in order, just like how cases overlapping is resolved in term-level
functions.
